A Logical Solution of the Linearity Challenge

### 2025-11-30 paper/Context.lagda.tex
co de bruijn

To conclude, we define a predicate $\AgdaDatatype{Un}~\Context$ that holds
whenever $\Context$ is \emph{unrestricted}, namely when it is made of types of
the form $?A$. We will need this predicate in the definition of a server, which
must obey the typing rule \ServerRule of \Cref{tab:typing-rules}.

@@

>>>>

To conclude the modeling of typing contexts, we define a predicate $\AgdaDatatype{Un}~\Context$ that holds
whenever $\Context$ is \emph{unrestricted}, namely when it is made of types of
the form $?A$. We will need this predicate in the definition of a server, which
must obey the typing rule \ServerRule of \Cref{tab:typing-rules}.

@@

It is a simple fact that the merging of two unrestricted contexts
$\Context_1$ and $\Context_2$ results in an unrestricted context.\Luca{Serve?}

\begin{code}
+-un : ∀{Γ Γ₁ Γ₂} → Γ ≃ Γ₁ + Γ₂ → Un Γ₁ → Un Γ₂ → Un Γ
\end{code}
\begin{code}[hide]
+-un • un-[] un-[] = un-[]
+-un (< p) (un-∷ un₁) un₂ = un-∷ (+-un p un₁ un₂)
+-un (> p) un₁ (un-∷ un₂) = un-∷ (+-un p un₁ un₂)
\end{code}

### 2025-11-30 paper/Permutations.lagda.tex
co de bruijn

We define typing context permutations inductively, as a binary relation
\AgdaDatatype{\_\#\_}:

\begin{code}
data _#_ : Context → Context → Set where
  #refl  : ∀{Γ} → Γ # Γ
  #here  : ∀{A B Γ} → (A ∷ B ∷ Γ) # (B ∷ A ∷ Γ)
  #next  : ∀{A Γ Δ} → Γ # Δ → (A ∷ Γ) # (A ∷ Δ)
  #tran  : ∀{Γ Δ Θ} → Γ # Δ → Δ # Θ → Γ # Θ
\end{code}

This definition is essentially the same as the notion of list permutation in
Agda's standard library. Each constructor represents a particular kind of
permutation: \AgdaInductiveConstructor{\#refl} for the trivial permutation that
does not change anything; \AgdaInductiveConstructor{\#here} for the permutation
that swaps the first two elements of a typing context;
\AgdaInductiveConstructor{\#next} for the permutation that applies a
sub-permutation to the tail of a typing context;
\AgdaInductiveConstructor{\#tran} for the sequential composition of
permutations.
%
We have preferred defining our own notion of permutation for simplicity: the
\AgdaInductiveConstructor{\#here} constructor does not need a sub-permutation
for the tail of the typing context, which can always be performed, if needed,
using \AgdaInductiveConstructor{\#next} and \AgdaInductiveConstructor{\#tran}.
Also, and more importantly, this definition is \emph{monomorphic} (it does not
need to operate on lists of arbitrary type), resulting in a more streamlined
usage in the rest of the code.

\begin{code}[hide]
#sym : ∀{Γ Δ} → Γ # Δ → Δ # Γ
#sym #refl = #refl
#sym #here = #here
#sym (#next π) = #next (#sym π)
#sym (#tran π π′) = #tran (#sym π′) (#sym π)

#empty-inv : ∀{Γ} → [] # Γ → Γ ≡ []
#empty-inv #refl = refl
#empty-inv (#tran π π′) rewrite #empty-inv π | #empty-inv π′ = refl

#singleton-inv : ∀{A Γ} → [ A ] # Γ → Γ ≡ [ A ]
#singleton-inv {Γ = Γ} #refl = refl
#singleton-inv {Γ = Γ} (#next π) rewrite #empty-inv π = refl
#singleton-inv {Γ = Γ} (#tran π π′) rewrite #singleton-inv π | #singleton-inv π′ = refl

#rot : ∀{A B C Γ} → (A ∷ B ∷ C ∷ Γ) # (C ∷ A ∷ B ∷ Γ)
#rot = #tran (#next #here) #here
\end{code}

Just like permutations from Agda's standard library, also \AgdaDatatype{\_\#\_}
enjoys a number of expected (and useful) properties, such as symmetry. Among all
properties, in the following we will make use of a specific one that connects
context splitting and permutations: if $\Splitting\ContextC\ContextD\ContextE$,
then $\ContextC$ is a permutation of the concatenation of $\ContextD$ and
$\ContextE$.\Luca{Ci sono molte altre proprietà non descritte, controllare
quelle che servono davvero.}

\begin{code}[hide]
#split : ∀{Γ Γ₁ Γ₂ Δ} → Γ # Δ → Γ ≃ Γ₁ + Γ₂ → ∃[ Δ₁ ] ∃[ Δ₂ ] (Δ ≃ Δ₁ + Δ₂ × Γ₁ # Δ₁ × Γ₂ # Δ₂)
#split #refl p = _ , _ , p , #refl , #refl
#split (#next π) (< p) with #split π p
... | Δ₁ , Δ₂ , q , π₁ , π₂ = _ ∷ Δ₁ , Δ₂ , < q , #next π₁ , π₂
#split (#next π) (> p) with #split π p
... | Δ₁ , Δ₂ , q , π₁ , π₂ = Δ₁ , _ ∷ Δ₂ , > q , π₁ , #next π₂
#split #here (< < p) = _ , _ , < < p , #here , #refl
#split #here (< > p) = _ , _ , > < p , #refl , #refl
#split #here (> < p) = _ , _ , < > p , #refl , #refl
#split #here (> > p) = _ , _ , > > p , #refl , #here
#split (#tran π π′) p with #split π p
... | Θ₁ , Θ₂ , p′ , π₁ , π₂ with #split π′ p′
... | Δ₁ , Δ₂ , q , π₁′ , π₂′ = Δ₁ , Δ₂ , q , #tran π₁ π₁′ , #tran π₂ π₂′

#one+ : ∀{A Γ Γ′ Δ} → Γ # Δ → Γ ≃ A , Γ′ → ∃[ Δ′ ] (Δ ≃ A , Δ′ × Γ′ # Δ′)
#one+ π p with #split π p
... | _ , _ , q , π₁ , π₂ rewrite #singleton-inv π₁ = _ , q , π₂

#shift : ∀{A Γ Δ} → (Γ ++ A ∷ Δ) # (A ∷ Γ ++ Δ)
#shift {_} {[]} = #refl
#shift {_} {B ∷ Γ} = #tran (#next #shift) #here
\end{code}

\begin{code}
#append : ∀{Γ Γ₁ Γ₂} → Γ ≃ Γ₁ + Γ₂ → (Γ₁ ++ Γ₂) # Γ
\end{code}
\begin{code}[hide]
#append • = #refl
#append (< p) = #next (#append p)
#append (> p) = #tran #shift (#next (#append p))

#left : ∀{Γ Δ Θ} → Γ # Δ → (Θ ++ Γ) # (Θ ++ Δ)
#left {Θ = []} π = π
#left {Θ = _ ∷ Θ} π = #next (#left π)

#un : ∀{Γ Δ} → Γ # Δ → Un Γ → Un Δ
#un #refl un = un
#un (#next π) (un-∷ un) = un-∷ (#un π un)
#un #here (un-∷ (un-∷ un)) = un-∷ (un-∷ un)
#un (#tran π π′) un = #un π′ (#un π un)

#un+ : ∀{Γ Γ₁ Γ₂} → Γ ≃ Γ₁ + Γ₂ → Un Γ₁ → Un Γ₂ → Un Γ
#un+ • un-[] un-[] = un-[]
#un+ (< p) (un-∷ un₁) un₂ = un-∷ (#un+ p un₁ un₂)
#un+ (> p) un₁ (un-∷ un₂) = un-∷ (#un+ p un₁ un₂)
\end{code}

>>>>

We define typing context permutations inductively, as a binary relation
\AgdaDatatype{\_↭\_}:

\begin{code}
data _↭_ : Context → Context → Set where
  refl   : ∀{Γ} → Γ ↭ Γ
  swap   : ∀{A B Γ} → (A ∷ B ∷ Γ) ↭ (B ∷ A ∷ Γ)
  prep   : ∀{A Γ Δ} → Γ ↭ Δ → (A ∷ Γ) ↭ (A ∷ Δ)
  trans  : ∀{Γ Δ Θ} → Γ ↭ Δ → Δ ↭ Θ → Γ ↭ Θ
\end{code}

This definition is essentially the same as the notion of list permutation in
Agda's standard library. Each constructor represents a particular kind of
permutation: \AgdaInductiveConstructor{refl} for the trivial permutation that
does not change anything; \AgdaInductiveConstructor{swap} for the permutation
that swaps the first two elements of a typing context;
\AgdaInductiveConstructor{prep} for the permutation that applies a
sub-permutation to the tail of a typing context;
\AgdaInductiveConstructor{trans} for the sequential composition of permutations.
%
We have preferred defining our own notion of permutation for simplicity: the
\AgdaInductiveConstructor{swap} constructor does not need a sub-permutation for
the tail of the typing context, which can always be performed, if needed, using
\AgdaInductiveConstructor{prep} and \AgdaInductiveConstructor{trans}. Also, and
more importantly, this definition is \emph{monomorphic} (it does not need to
operate on lists of arbitrary type), resulting in a more streamlined usage in
the rest of the code.

\begin{code}[hide]
↭sym : ∀{Γ Δ} → Γ ↭ Δ → Δ ↭ Γ
↭sym refl = refl
↭sym swap = swap
↭sym (prep π) = prep (↭sym π)
↭sym (trans π π′) = trans (↭sym π′) (↭sym π)

↭empty-inv : ∀{Γ} → [] ↭ Γ → Γ ≡ []
↭empty-inv refl = refl
↭empty-inv (trans π π′) rewrite ↭empty-inv π | ↭empty-inv π′ = refl

↭solo-inv : ∀{A Γ} → [ A ] ↭ Γ → Γ ≡ [ A ]
↭solo-inv refl = refl
↭solo-inv (prep π) rewrite ↭empty-inv π = refl
↭solo-inv (trans π π′) rewrite ↭solo-inv π | ↭solo-inv π′ = refl

↭rot : ∀{A B C Γ} → (A ∷ B ∷ C ∷ Γ) ↭ (C ∷ A ∷ B ∷ Γ)
↭rot = trans (prep swap) swap
\end{code}

Just like permutations from Agda's standard library, also \AgdaDatatype{\_↭\_}
enjoys a number of expected (and useful) properties, such as symmetry. Among all
properties, in the following we will make use of a specific one that connects
context splitting and permutations: if $\Splitting\ContextC\ContextD\ContextE$,
then $\ContextC$ is a permutation of the concatenation of $\ContextD$ and
$\ContextE$.\Luca{Ci sono molte altre proprietà non descritte, controllare
quelle che servono davvero.}

\begin{code}[hide]
↭split : ∀{Γ Γ₁ Γ₂ Δ} → Γ ↭ Δ → Γ ≃ Γ₁ + Γ₂ → ∃[ Δ₁ ] ∃[ Δ₂ ] (Δ ≃ Δ₁ + Δ₂ × Γ₁ ↭ Δ₁ × Γ₂ ↭ Δ₂)
↭split refl p = _ , _ , p , refl , refl
↭split (prep π) (< p) with ↭split π p
... | Δ₁ , Δ₂ , q , π₁ , π₂ = _ ∷ Δ₁ , Δ₂ , < q , prep π₁ , π₂
↭split (prep π) (> p) with ↭split π p
... | Δ₁ , Δ₂ , q , π₁ , π₂ = Δ₁ , _ ∷ Δ₂ , > q , π₁ , prep π₂
↭split swap (< < p) = _ , _ , < < p , swap , refl
↭split swap (< > p) = _ , _ , > < p , refl , refl
↭split swap (> < p) = _ , _ , < > p , refl , refl
↭split swap (> > p) = _ , _ , > > p , refl , swap
↭split (trans π π′) p with ↭split π p
... | Θ₁ , Θ₂ , p′ , π₁ , π₂ with ↭split π′ p′
... | Δ₁ , Δ₂ , q , π₁′ , π₂′ = Δ₁ , Δ₂ , q , trans π₁ π₁′ , trans π₂ π₂′

↭solo : ∀{A Γ Γ′ Δ} → Γ ↭ Δ → Γ ≃ A , Γ′ → ∃[ Δ′ ] (Δ ≃ A , Δ′ × Γ′ ↭ Δ′)
↭solo π p with ↭split π p
... | _ , _ , q , π₁ , π₂ rewrite ↭solo-inv π₁ = _ , q , π₂

↭shift : ∀{A Γ Δ} → (Γ ++ A ∷ Δ) ↭ (A ∷ Γ ++ Δ)
↭shift {_} {[]} = refl
↭shift {_} {_ ∷ _} = trans (prep ↭shift) swap
\end{code}

\begin{code}
↭concat : ∀{Γ Γ₁ Γ₂} → Γ ≃ Γ₁ + Γ₂ → (Γ₁ ++ Γ₂) ↭ Γ
\end{code}
\begin{code}[hide]
↭concat • = refl
↭concat (< p) = prep (↭concat p)
↭concat (> p) = trans ↭shift (prep (↭concat p))

↭left : ∀{Γ Δ Θ} → Γ ↭ Δ → (Θ ++ Γ) ↭ (Θ ++ Δ)
↭left {Θ = []} π = π
↭left {Θ = _ ∷ _} π = prep (↭left π)

↭un : ∀{Γ Δ} → Γ ↭ Δ → Un Γ → Un Δ
↭un refl un = un
↭un (prep π) (un-∷ un) = un-∷ (↭un π un)
↭un swap (un-∷ (un-∷ un)) = un-∷ (un-∷ un)
↭un (trans π π′) un = ↭un π′ (↭un π un)
\end{code}

### 2025-11-30 paper/Process.lagda.tex
co de bruijn

We adopt an \emph{intrinsically typed} representation of processes, so that a
process is also a proof derivation showing that the process is well typed. This
choice increases the effort in the definition of the datatypes for processes and
their operational semantics, but pays off in the rest of the formalization for a
number of reasons:

\begin{itemize}
\item channel names are encoded in the context splitting proofs occurring within
  processes, hence they do not need to be represented explicitly.
\item processes and typing rules are conflated in the same datatype, thus
  reducing the overall number of datatypes in the formalization;
\item the typing preservation results are embedded in the definition of the
  operational semantics of processes and do not need to be proved separately.
\end{itemize}

We now illustrate the definition of the \AgdaDatatype{Process} datatype and
describe each constructor in detail. The datatype is indexed by a typing
context:

\begin{code}
data Process : Context → Set where
  link      : ∀{A Γ} → Γ ≃ [ A ] + [ dual A ] → Process Γ
  fail      : ∀{Γ Δ} → Γ ≃ ⊤ , Δ → Process Γ
  wait      : ∀{Γ Δ} → Γ ≃ ⊥ , Δ → Process Δ → Process Γ
  close     : Process [ 𝟙 ]
  case      : ∀{A B Γ Δ} → Γ ≃ A & B , Δ →
              Process (A ∷ Δ) → Process (B ∷ Δ) → Process Γ
  select    : ∀{A B Γ Δ} (x : Bool) → Γ ≃ A ⊕ B , Δ →
              Process ((if x then A else B) ∷ Δ) → Process Γ
  join      : ∀{A B Γ Δ} → Γ ≃ A ⅋ B , Δ → Process (B ∷ A ∷ Δ) → Process Γ
  fork      : ∀{A B Γ Δ Γ₁ Γ₂} → Γ ≃ A ⊗ B , Δ → Δ ≃ Γ₁ + Γ₂ →
              Process (A ∷ Γ₁) → Process (B ∷ Γ₂) → Process Γ
  all       : ∀{A Γ Δ} → Γ ≃ `∀ A , Δ →
              ((B : Type) → Process (subst [ B /_] A ∷ Δ)) → Process Γ
  ex        : ∀{A B Γ Δ} → Γ ≃ `∃ A , Δ → Process (subst [ B /_] A ∷ Δ) → Process Γ
  server    : ∀{A Γ Δ} → Γ ≃ `! A , Δ → Un Δ → Process (A ∷ Δ) → Process Γ
  client    : ∀{A Γ Δ} → Γ ≃ `? A , Δ → Process (A ∷ Δ) → Process Γ
  weaken    : ∀{A Γ Δ} → Γ ≃ `? A , Δ → Process Δ → Process Γ
  contract  : ∀{A Γ Δ} → Γ ≃ `? A , Δ → Process (`? A ∷ `? A ∷ Δ) → Process Γ
  cut       : ∀{A Γ Γ₁ Γ₂} → Γ ≃ Γ₁ + Γ₂ →
              Process (A ∷ Γ₁) → Process (dual A ∷ Γ₂) → Process Γ
\end{code}

@@

\begin{code}
#process : ∀{Γ Δ} → Γ # Δ → Process Γ → Process Δ
\end{code}
\begin{code}[hide]
#process π (link p) with #one+ π p
... | Δ′ , q , π′ with #singleton-inv π′
... | refl = link q
#process π (fail p) with #one+ π p
... | Δ′ , q , π′ = fail q
#process π (wait p P) with #one+ π p
... | Δ′ , q , π′ = wait q (#process π′ P)
#process π close with #singleton-inv π
... | refl = close
#process π (case p P Q) with #one+ π p
... | Δ′ , q , π′ = case q (#process (#next π′) P) (#process (#next π′) Q)
#process π (select x p P) with #one+ π p
... | Δ′ , q , π′ = select x q (#process (#next π′) P)
#process π (join p P) with #one+ π p
... | Δ′ , q , π′ = join q (#process (#next (#next π′)) P)
#process π (fork p q P Q) with #one+ π p
... | Δ′ , p′ , π′ with #split π′ q
... | Δ₁ , Δ₂ , q′ , π₁ , π₂ = fork p′ q′ (#process (#next π₁) P) (#process (#next π₂) Q)
#process π (all p P) with #one+ π p
... | Δ' , q , π' = all q λ B → #process (#next π') (P B)
#process π (ex p P) with #one+ π p
... | Δ' , q , π' = ex q (#process (#next π') P)
#process π (server p un P) with #one+ π p
... | Δ′ , q , π′ = server q (#un π′ un) (#process (#next π′) P)
#process π (client p P) with #one+ π p
... | Δ′ , q , π′ = client q (#process (#next π′) P)
#process π (weaken p P) with #one+ π p
... | Δ′ , q , π′ = weaken q (#process π′ P)
#process π (contract p P) with #one+ π p
... | Δ′ , q , π′ = contract q (#process (#next (#next π′)) P)
#process π (cut p P Q) with #split π p
... | Δ₁ , Δ₂ , q , π₁ , π₂ = cut q (#process (#next π₁) P) (#process (#next π₂) Q)
\end{code}

>>>>

We adopt a \emph{nameless, intrinsically typed} representation of processes,
that is: a channel is identified simply by the presence of its type in the
typing context; a process is also a proof derivation showing that the process is
well typed. This choice increases the effort in the definition of the datatypes
for processes and their operational semantics, but pays off in the rest of the
formalization for a number of reasons:
\begin{itemize}
\item we use context splitting not only to separate resources used by parallel
  processes (\cf \ForkRule and \CutRule), but also to witness that a channel is
  present in the typing context;
\item we conflate processes and typing rules in the same datatype, thus reducing
  the overall number of datatypes defined in the formalization;
\item we embed the typing preservation results within the very definition of the
  operational semantics of processes instead of proving them separately
  (\Cref{sec:agda-semantics}).
\end{itemize}

Our nameless modeling of channels is known in the literature as the \emph{co-de
Bruijn} representation~\cite{McBride18}. The proof that the type of a channel is
found within a typing context essentially encodes the de Bruijn index of the
channel in the typing context of a given process. However, the typing context
only contains types for those channels that actually occur in the process.
Context splitting makes sure that channels are appropriately (and above all
linearly) distributed among sub-processes, so that each channel occurs (and is
used) exactly once.\footnote{Strictly speaking, the typing context of a
\AgdaInductiveConstructor{fail} process may contain channels that do not occur
(and therefore are not used) by the process, but we consider this case as an
exception that does not alter in a substantial way the provided intuition
underlying the use of co-de Bruijn representation.}

We now illustrate the definition of the \AgdaDatatype{Process} datatype and
describe its key aspects in detail. The datatype is indexed by a typing context:

\begin{code}
data Process : Context → Set where
  link      : ∀{A Γ} → Γ ≃ A , [ dual A ] → Process Γ
  fail      : ∀{Γ Δ} → Γ ≃ ⊤ , Δ → Process Γ
  wait      : ∀{Γ Δ} → Γ ≃ ⊥ , Δ → Process Δ → Process Γ
  close     : Process [ 𝟙 ]
  case      : ∀{A B Γ Δ} → Γ ≃ A & B , Δ →
              Process (A ∷ Δ) → Process (B ∷ Δ) → Process Γ
  select    : ∀{A B Γ Δ} (x : Bool) → Γ ≃ A ⊕ B , Δ →
              Process ((if x then A else B) ∷ Δ) → Process Γ
  join      : ∀{A B Γ Δ} → Γ ≃ A ⅋ B , Δ → Process (B ∷ A ∷ Δ) → Process Γ
  fork      : ∀{A B Γ Δ Γ₁ Γ₂} → Γ ≃ A ⊗ B , Δ → Δ ≃ Γ₁ + Γ₂ →
              Process (A ∷ Γ₁) → Process (B ∷ Γ₂) → Process Γ
  all       : ∀{A Γ Δ} → Γ ≃ `∀ A , Δ →
              ((B : Type) → Process (subst [ B /_] A ∷ Δ)) → Process Γ
  ex        : ∀{A B Γ Δ} → Γ ≃ `∃ A , Δ → Process (subst [ B /_] A ∷ Δ) → Process Γ
  server    : ∀{A Γ Δ} → Γ ≃ `! A , Δ → Un Δ → Process (A ∷ Δ) → Process Γ
  client    : ∀{A Γ Δ} → Γ ≃ `? A , Δ → Process (A ∷ Δ) → Process Γ
  weaken    : ∀{A Γ Δ} → Γ ≃ `? A , Δ → Process Δ → Process Γ
  contract  : ∀{A Γ Δ} → Γ ≃ `? A , Δ → Process (`? A ∷ `? A ∷ Δ) → Process Γ
  cut       : ∀{A Γ Γ₁ Γ₂} → Γ ≃ Γ₁ + Γ₂ →
              Process (A ∷ Γ₁) → Process (dual A ∷ Γ₂) → Process Γ
\end{code}

@@

\begin{code}
↭process : ∀{Γ Δ} → Γ ↭ Δ → Process Γ → Process Δ
\end{code}
\begin{code}[hide]
↭process π (link p) with ↭solo π p
... | Δ′ , q , π′ rewrite ↭solo-inv π′ = link q
↭process π (fail p) with ↭solo π p
... | Δ′ , q , π′ = fail q
↭process π (wait p P) with ↭solo π p
... | Δ′ , q , π′ = wait q (↭process π′ P)
↭process π close rewrite ↭solo-inv π = close
↭process π (case p P Q) with ↭solo π p
... | Δ′ , q , π′ = case q (↭process (prep π′) P) (↭process (prep π′) Q)
↭process π (select x p P) with ↭solo π p
... | Δ′ , q , π′ = select x q (↭process (prep π′) P)
↭process π (join p P) with ↭solo π p
... | Δ′ , q , π′ = join q (↭process (prep (prep π′)) P)
↭process π (fork p q P Q) with ↭solo π p
... | Δ′ , p′ , π′ with ↭split π′ q
... | Δ₁ , Δ₂ , q′ , π₁ , π₂ = fork p′ q′ (↭process (prep π₁) P) (↭process (prep π₂) Q)
↭process π (all p P) with ↭solo π p
... | Δ' , q , π' = all q λ B → ↭process (prep π') (P B)
↭process π (ex p P) with ↭solo π p
... | Δ' , q , π' = ex q (↭process (prep π') P)
↭process π (server p un P) with ↭solo π p
... | Δ′ , q , π′ = server q (↭un π′ un) (↭process (prep π′) P)
↭process π (client p P) with ↭solo π p
... | Δ′ , q , π′ = client q (↭process (prep π′) P)
↭process π (weaken p P) with ↭solo π p
... | Δ′ , q , π′ = weaken q (↭process π′ P)
↭process π (contract p P) with ↭solo π p
... | Δ′ , q , π′ = contract q (↭process (prep (prep π′)) P)
↭process π (cut p P Q) with ↭split π p
... | Δ₁ , Δ₂ , q , π₁ , π₂ = cut q (↭process (prep π₁) P) (↭process (prep π₂) Q)
\end{code}
